---
title: 'Tutorial 4: Advanced Visualization'
runtime: shiny_prerendered
output: learnr::tutorial
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
load("601_AdvancedVisualization.rdata")
tutorial_options(exercise.timelimit = 10)
```

## **Introduction**

In this tutorial, we will build on what we learned in our Intro to Visualization tutorial. Now, we will look to add dimensions to our plot, analyze differences across groups, and think about how best to design our graphics. All of this continues to build on the core idea that visualization is central to data science.

## **GGPlot: Facet_Wrap** 

One really useful addition we can make to `ggplot()` is `facet_wrap()`. Facet wrapping will divide the plot into subplots based on different values of another discrete variable in the dataset, kind of like `group_by()` with graphs.

There are multiple components within `facet_wrap()`. First is `vars()`, in which you choose the variable to base subplot divisions upon.

In the example below, we divide the histogram of **p50_earnings** into subplots based on **year_postgrad** with the addition of: `facet_wrap(vars(year_postgrad))`

```{r, echo=TRUE}
#Divide the p50_earnings graph below based on year_postgrad
ggplot(pseo, aes(p50_earnings)) + 
  geom_histogram(binwidth = 20000) + 
  labs(title = "50th Percentile Earnings") + 
  theme_bw() +
  facet_wrap(vars(year_postgrad))
```

Also, take note that we changed the `binwidth` of `geom_histogram()`. This just changes the width of the bars in the histogram. You can experiment with different values on your own or choose to leave it at default.

These three graphs all have the same scales for their x and y axes, even though the range of values are quite different in each. You can change the scales using `scales =` within `facet_wrap()`. Setting `scales = "free"` will adjust both the x and y axis for all graphs, or you can exclusively adjust the x axis with `scales = "free_x"` or the y axis with `scales = "free_y"`. See what this means in the examples below. 

When `scales = "free"`:

```{r, echo=TRUE}
ggplot(pseo, aes(p50_earnings)) + 
  geom_histogram(binwidth = 20000) + 
  labs(title = "50th Percentile Earnings") + 
  theme_bw() +
  facet_wrap(vars(year_postgrad), scales = "free")
```

When `scales = "free_x"`:

```{r, echo=TRUE}
ggplot(pseo, aes(p50_earnings)) + 
  geom_histogram(binwidth = 20000) + 
  labs(title = "50th Percentile Earnings") + 
  theme_bw() +
  facet_wrap(vars(year_postgrad), scales = "free_x")
```

When `scales = "free_y"`:

```{r, echo=TRUE}
ggplot(pseo, aes(p50_earnings)) + 
  geom_histogram(binwidth = 20000) + 
  labs(title = "50th Percentile Earnings") + 
  theme_bw() +
  facet_wrap(vars(year_postgrad), scales = "free_y")
```

Pay attention to the axes to see what the differences are between these. 

### Exercise 6

Take the plot of presidential vote answers.

```{r, echo=TRUE}
ggplot(anes, aes(Vote_16)) + geom_bar() +
  theme_bw() +
  labs(title = "ANES Survey of 2016 Presidential Vote", y = "Number of Respondents", x = "Presidential Vote")
```

Add a `facet_wrap()` function that divides the plot based on the variable **Class**. See what axis scaling option makes the most sense.

```{r facet, exercise=TRUE}
#Divide the graph below based on the variable Class
ggplot(anes, aes(Vote_16)) + geom_bar() +
  theme_bw() +
  labs(title = "ANES Survey of 2016 Presidential Vote", y = "Number of Respondents", x = "Presidential Vote")
```

```{r facet-solution}
#Divide the graph below based on the variable Class
ggplot(anes, aes(Vote_16)) + geom_bar() +
  theme_bw() +
  labs(title = "ANES Survey of 2016 Presidential Vote", y = "Number of Respondents", x = "Presidential Vote") +
  facet_wrap(vars(Class), scale = "free")
```

## **GGPlot: Fill**

One last modification to learn within `ggplot()` is `fill`, something we briefly discussed earlier. This setting is made within the base `aes()` function, so that you can list another variable to divide individual points/bars into different colors.  

In the example below, the histogram of **p50_earnings** is filled based on the variable **deglevl**. 

```{r, echo=TRUE}
#Fill the p50_earnings graph based on the variable deglevl
ggplot(pseo, aes(p50_earnings, fill = deglevl)) + 
  geom_histogram(binwidth = 20000) + 
  labs(title = "50th Percentile Earnings") + 
  theme_bw() +
  facet_wrap(vars(year_postgrad), scales = "free_x")
```

As you can see, there are now multiple colors as well as a color code that is created on the right. 

We have now learned how to create a clean-looking, interesting graph with `ggplot()`. 

### Exercise 7

Take the plot of presidential vote answers.

```{r, echo=TRUE}
ggplot(anes, aes(Vote_16)) + geom_bar() +
  theme_bw() +
  labs(title = "ANES Survey of 2016 Presidential Vote", y = "Number of Respondents", x = "Presidential Vote")
```

Modify this graph with `fill` so that it is color coded based on the variable **Class**. 

```{r fill, exercise=TRUE}
#Fill graph below based on the variable Class
ggplot(anes, aes(Vote_16)) + geom_bar() +
  theme_bw() +
  labs(title = "ANES Survey of 2016 Presidential Vote", y = "Number of Respondents", x = "Presidential Vote")
```

```{r fill-solution}
#Fill graph below based on the variable Class
ggplot(anes, aes(Vote_16, fill = Class)) + geom_bar() +
  theme_bw() +
  labs(title = "ANES Survey of 2016 Presidential Vote", y = "Number of Respondents", x = "Presidential Vote") 
```

## **Wrangling and Plotting**

The plots so far have been good, but still not ideal. Sometimes your graphs are limited by the data available to you and its format. To overcome these limitations, we can incorporate data wrangling functions from Tutorial 2 into our `ggplot()` graphs.

We will look at two different examples to incorporate `filter()` and `gather()` into plotting.

#### **Filter**

As you probably saw in the last exercises, the bar graphs of 2016 presidential vote are largely skewed by `NA` answers. It doesn't look great.

```{r echo = TRUE}
ggplot(anes, aes(Vote_16, fill = Class)) + geom_bar() +
  theme_bw() +
  labs(title = "ANES Survey of 2016 Presidential Vote", y = "Number of Respondents", x = "Presidential Vote") 
```

Although statistical tests will ignore `NA` values in calculations, `ggplot` will graph them like any other value. Fortunately, we can counteract this.

We can use `filter()` to hide `NA` values from `ggplot()`. In the example below, we use piping (`%>%`) to filter out `NA` values before plotting them. We'll come back to piping soon. For now, you just need to know that because we are piping, the function begins with `anes %>%` and then we can leave `anes` out of the beginning of `ggplot()`. The logical operator `!` and the function `is.na()` are used together to select rows where Class is NOT `NA`. See it in action.

```{r echo = TRUE}
anes %>%
  filter(! is.na(Class)) %>%
  filter(! is.na(Vote_16)) %>%
  ggplot(aes(Vote_16, fill = Class)) + geom_bar() +
  theme_bw() +
  labs(title = "ANES Survey of 2016 Presidential Vote", y = "Number of Respondents", x = "Presidential Vote") 
```

Much better.

#### **Gather**

The structure of the `electric` dataset is rather odd; each column is a separate year. Recall that we discussed this in our second tutorial. What would make more sense is to have a single column that is called **year** and another column called **electric**. To do this in R, we can use `gather()`. This setting offers a great opportunity to think about why this is important; without using `gather()`, we can only make scatterplots of electricity access values from one year compared to another year. This graph is not very helpful:

```{r echo = TRUE}
ggplot(electric, aes(`1990`, `2016`)) + geom_point() +
theme_bw()
```

However, if we use `gather()` to turn **year** and **access** into columns, we can make a scatterplot that compares access over time. This again uses piping (`%>%`) which is code we'll discuss at greater length later on. The general idea, though, is that you are "passing" an object through a series of functions. So here we pipe the `electric` dataset through a gather function by begininning our code with `electric %>%`, then following that with `gather()`. 

```{r echo = TRUE}
electric %>%
  gather(`1990`:`2016`, key = "year", value = "access") %>%
  ggplot(aes(year, access)) + 
  geom_point() + 
  labs(title = "World Electricity Access 1990-2016") + 
  theme_bw()
```

Well, this still looks confusing. Now, we will use `filter()` to select a specific country to observe. Below, we examine electricity access in Iraq and finally make an insightful bivariate plot from the **electric** data. 

```{r echo = TRUE}
electric %>%
  gather(`1990`:`2016`, key = "year", value = "access") %>%
  filter(`Country Name` == "Iraq") %>%
  ggplot(aes(year, access)) + 
  geom_point() + 
  labs(title = "Iraq Electricity Access 1990-2016") + 
  theme_bw()
```

The y axis will automatically adjust to the specific values for Iraq, as well.

### Exercise 8

Try using `gather()` and `ggplot()` to graph the electricity access of all countries over time, but use `geom_boxplot()` as your geometry. There should be a boxplot for every year.

```{r wrangle, exercise=TRUE}
#Make a series of boxplots comparing electricity access over time

```

```{r wrangle-solution}
#Make a series of boxplots comparing electricity access over time
electric %>%
  gather(`1990`:`2016`, key = "year", value = "access") %>%
  ggplot(aes(year, access)) + 
  geom_boxplot() + 
  labs(title = "World Electricity Access 1990-2016") + 
  theme_bw()
```

## **Conclusion & Glossary**

You now know how to use `ggplot()` to create basic and advanced graphs for all types of variables and adjust various settings. You are also prepared to wrangle messy data in order to prepare it for graphing. In truth, you have only scratched the surface of what is possible with `ggplot()`. [Check out this cheatsheet to see all the available geometries, scales, themes and labels that can be used.](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf) However, what we have covered in these tutorials should get you pretty far.

Coming up in the next tutorial... we will turn back to our datasets, and work to transform the variables.

#### Function Glossary

Here's a recap of the functions you have encountered so far:

**Tutorial 1: R Basics and Assigning Variables**

* Basic mathematics: `+`,`-`,`*`,`/`
* Logical operators: `==`,`>`,`<`,`>=`,`>=`
* Assign name: `<-`
* Create vector: `c()`

**Tutorial 2: Reading and Describing Data**

* Preview data: `head()`
* Get dimensions: `dim()`
* Get column names and numbers: `colnames()`
* Select specific columns: `select()`
  + `starts_with()`
  + `ends_with()`
  + `contains()`
* Get values of column: `table()`
* Get proportions of values of column: `prop.table()`
* Collapse multiple columns into new rows: `gather()`
* Turn row data into new columns: `spread()`

**Tutorial 3: Intro to Visualization**

* Make crosstabs: `xtabs()`
* Make graphs: `ggplot()`
  + Histogram: `geom_histogram()`
  + Density: geom_density()
  + Bar graph: `geom_bar()`
  + Scatterplot: `geom_point()`
  + Boxplot: `geom_boxplot()`
  + Violin plot: `geom_violin()`
  + Smooth line plot: `geom_smooth()`

**Tutorial 4: Advanced Visualization**

* Make crosstabs: `xtabs()`
* Make graphs: `ggplot()`
  + Divide plot into subplots: `facet_wrap()`
* Color observations: `aes(fill =)`

  